---
title: "Morphometrics"
author: "Brian C. O'Meara"
date: "6/8/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ape)
library(abind)
library(phytools)
library(geiger)
```

First, need to make a few plotting functions:

```{r, echo=FALSE}
#' Plot outlines of taxa forms
#' @param traits Matrix of measurements; rownames are taxa, cols are traits
#' @param n Number of taxa in matrix to plot
#' @param color Color to use
PlotForms <- function(traits, n=NULL, color=rgb(0,0,0,.5)) {
  x.values <- unlist(traits[seq(from=1, to=ncol(traits), by=2),])
  y.values <- unlist(traits[seq(from=2, to=ncol(traits), by=2),])
  plot(x.values, y.values, type="n", bty="n",asp=1)
  if(is.null(n)) {
    n=nrow(traits) 
  }
   for (taxon.index in sequence(n)) {
      PlotSingleForm(traits[taxon.index,], color=color) 
   }
}

#' Plot a single taxon
#' @param trait.vector A vector of values: x is odd entries, y are even
#' @param color Color of lines
PlotSingleForm <- function(trait.vector, color=rgb(0,0,0,.5)) {
  x.values <- unlist(trait.vector[seq(from=1, to=length(trait.vector), by=2)])
  y.values <- unlist(trait.vector[seq(from=2, to=length(trait.vector), by=2)])
  #lines(point.matrix$x, point.matrix$y, col=color)
  polygon(x.values, y.values, border=color)
}

#' Plot outlines of taxa forms
#' @param traits Matrix of measurements; rownames are taxa, cols are traits
#' @param pc Principal components to plot
#' @param just.mean If TRUE, plots the mean points; if FALSE, overplots all the fish
PlotMeanAndPC <- function(traits, pc=NULL, just.mean=TRUE) {
  x.values <- unlist(traits[seq(from=1, to=ncol(traits), by=2),])
  y.values <- unlist(traits[seq(from=2, to=ncol(traits), by=2),])
  plot(x.values, y.values, type="n", bty="n",asp=1)
  all.means <- apply(traits, 2, mean)
  x.means <- all.means[seq(from=1, to=ncol(traits), by=2)]
  y.means <- all.means[seq(from=2, to=ncol(traits), by=2)]
  x.pc <- pc[seq(from=1, to=ncol(traits), by=2)]
  y.pc <- pc[seq(from=2, to=ncol(traits), by=2)]
  if(just.mean) {
    PlotSingleForm(all.means, color=rgb(0,0,0,.5)) 
  } else {
     PlotForms(traits, color=rgb(0,0,0,.05))
     PlotSingleForm(all.means, color="black") 
  }
  if(!is.null(pc)) {
    for (i in sequence(length(x.pc))) {
      arrows(x0=x.means[i], y0=y.means[i], x1=x.means[i]+x.pc[i], y1=y.means[i]+y.pc[i],  length = 0.05, angle = 20, col = "red", lwd = 4)
    }
  }
}
```

Get the data into R

```{r}
phy <- ape::read.tree("data/Ex5-1/tree.tre")
phy <- ape::reorder.phylo(phy)
```

The ape package doesn't check to make sure taxa match between tree and data. We use treedata to check this, but could use functions in treeplyr instead.

Note the repeated lines here: would be better done in a loop, but I wanted to avoid having assignment of object names.

```{r}
data1 <- geiger::treedata(phy, read.table("data/Ex5-1/dataset1.txt", row.names=1), sort=TRUE)$data
data2 <- geiger::treedata(phy, read.table("data/Ex5-1/dataset2.txt", row.names=1), sort=TRUE)$data
data3 <- geiger::treedata(phy, read.table("data/Ex5-1/dataset3.txt", row.names=1), sort=TRUE)$data
data4 <- geiger::treedata(phy, read.table("data/Ex5-1/dataset4.txt", row.names=1), sort=TRUE)$data
data5 <- geiger::treedata(phy, read.table("data/Ex5-1/dataset5.txt", row.names=1), sort=TRUE)$data

phy <- geiger::treedata(phy, data1, sort=TRUE)$phy #and do the same for the phylo object

all.data <- abind::abind(data1, data2, data3, data4, data5, along=3) #makes a 3d array.
```

Look at the data
```{r}
PlotForms(data1)
```


We can get phylogenetic independent contrasts 

```{r}
pic.scaled <- apply(data1, 2, ape::pic, phy=phy, scaled=TRUE)
```

stop("How do you expect size of data1 and pic.scaled to compare? The first dataset has 100 taxa and 20 chars.")

```{r}
print("Original dataset size")
print(dim(data1))

print("PIC dataset")
print(dim(pic.scaled))
```

stop("What has changed?")

Let's focus on a couple of traits in more detail, just the first and second.

```{r}
x.trait <- 1
y.trait <- 2
```

Look at them on a tree

```{r}
phytools::phylomorphospace(phy, data1[,c(x.trait,y.trait)])
```


Look at correlations. How do they differ?

```{r, echo=FALSE}
plot(data1[,x.trait], data1[,y.trait], xlab=colnames(data1)[x.trait], ylab=colnames(data1)[y.trait], bty="n", pch=20, col=rgb(0,0,0,.5), main="Raw species traits")
lm.result <- lm(paste(colnames(data1)[y.trait],"~",colnames(data1)[x.trait], sep=" "), data=data.frame(data1))
abline(lm.result)

plot(pic.scaled[,x.trait], pic.scaled[,y.trait], xlab=colnames(pic.scaled)[x.trait], ylab=colnames(pic.scaled)[y.trait], bty="n", pch=20, col=rgb(0,0,0,.5), main="Contrasts")
lm.result <- lm(paste(colnames(pic.scaled)[y.trait],"~",colnames(pic.scaled)[x.trait], "- 1", sep=" "), data=data.frame(pic.scaled))
abline(lm.result)
#correlation moves from positive to negative, when you add the context of Contrast
```

We can look at PIC for a variety of traits. The trait itself is plotted on the diagonal.

```{r, echo=FALSE}
par(mfcol=c(min(5, ncol(pic.scaled)), min(5, ncol(pic.scaled))))
for (i in sequence(min(5, ncol(pic.scaled)))) {
	for (j in sequence(min(5, ncol(pic.scaled)))) {
		if (i!=j) {
			x.trait <- i
			y.trait <- j
			plot(pic.scaled[,x.trait], pic.scaled[,y.trait], xlab=colnames(pic.scaled)[x.trait], ylab=colnames(pic.scaled)[y.trait], bty="n", pch=20, col=rgb(0,0,0,.5))
			lm.result <- lm(paste(colnames(pic.scaled)[y.trait],"~",colnames(pic.scaled)[x.trait], "- 1", sep=" "), data=data.frame(pic.scaled))
			abline(lm.result)
		} else {
			contMap(phy, data1[,i])
		}
	}
}

#correlation and tree varied traits with PIC - "gut check" 
#each x is a trait, each y is a trait
```

Simulate some uncorrelated and correlated evolution on a phylogeny
```{r, echo=FALSE}
sim.phy <- compute.brlen(ape::stree(2^7, type="balanced"))
plot(sim.phy, show.tip.label=FALSE)
sim.traits <- geiger::sim.char(sim.phy, par=matrix(c(1, 0.0, 0.0, 1), nrow=2, byrow=TRUE), nsim=1, model="BM")[,,1]
par(mfcol=c(1,2))
plot(sim.traits[,1], sim.traits[,2], bty="n", main="Raw (truth = uncorrelated)")
pic1 <- pic(sim.traits[,1], phy=sim.phy)
pic2 <- pic(sim.traits[,2], phy=sim.phy)
plot(pic1, pic2, bty="n", main="PIC (truth = uncorrelated)")

sim.traits <- geiger::sim.char(sim.phy, par=matrix(c(1, 0.5, 0.5, 1), nrow=2, byrow=TRUE), nsim=1, model="BM")[,,1]
par(mfcol=c(1,2))
plot(sim.traits[,1], sim.traits[,2], bty="n", main="Raw (truth = correlated)")
pic1 <- pic(sim.traits[,1], phy=sim.phy)
pic2 <- pic(sim.traits[,2], phy=sim.phy)
plot(pic1, pic2, bty="n", main="PIC (truth = correlated)")

```


Now, an exercise: plot the mean with arrows showing the principal components.

```{r}
PC <- runif(ncol(data1)) #just making them up, REPLACE THIS
PlotMeanAndPC(data1, PC, just.mean=FALSE)
```

Other things to investigate:

* The same species in different simulations (use all.data: indices are [species, trait, replicate])
* Variance within species.
* Procrustes transformations

Notes:
datasets 1-5: independent simulations on shared tree
Joe - simulated with variances and covariances among traits
Need not be morphometric
May need to log-transform your data

